---
title: "Class 05: Data Visualization with GGPLOT"
author: "Libby Gilmore, pid: A69047570"
format: pdf
editor: visual
toc: true
---

### q1

For which phases is data visualization important in our scientific workflows? - all of the above

### q2

True or False? The ggplot2 package comes already installed with R? - FALSE

### q3

Which plot types are typically NOT used to compare distributions of numeric variables? - Network graphs

### q4

Which statement about data visualization with ggplot2 is incorrect? - ggplot2 is the only way to create plots in R

### creating scatter plots

```{r}
# install.packages("ggplot2")
library(ggplot2)

ggplot(cars) # this just plots an empty grey screen
```

```{r}
# specific aesthetic mappings with aes()
ggplot(cars) +
  aes(x=speed, y=dist)
```

#### Specify a geom layer with `geom_point()`

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```

#### q5

Which geometric layer should be used to create scatter plots in ggplot2? - geom_point()

#### q6

In your own RStudio can you add a trend line layer to help show the relationship between the plot variables with the geom_smooth() function?

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() + 
  geom_smooth()
```

#### q7

Argue with geom_smooth() to add a straight line from a linear model without the shaded standard error region?

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE)
```

#### q8

finish this plot by adding various label annotations with the labs() function and changing the plot look to a more conservative “black & white” theme by adding the theme_bw() function:

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(x="speed (mph)", y = "stopping distance (ft)",
       title = "Speed and Stopping Distances of Cars",
       caption = "dataset: 'cars' ") +
  theme_bw()
```

### Adding more plot aesthetics through `aes()`

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
#Q. Use the nrow() function to find out how many genes are in this dataset. What is your answer? 
nrow(genes)


#Q. Use the colnames() function and the ncol() function on the genes data frame to find out what the column names are (we will need these later) and how many columns there are. How many columns did you find? 
colnames(genes)

ncol(genes)


#Q. Use the table() function on the State column of this data.frame to find out how many ‘up’ regulated genes there are. What is your answer? 
table(genes$State)
        
#Q. Using your values above and 2 significant figures. What fraction of total genes is up-regulated in this dataset? 
round( table(genes$State)/nrow(genes) * 100, 2 )

```

```{r}
# Q. Complete the code below to produce the following plot

ggplot(genes) + 
    aes(x=Condition1, y=Condition2) +
    geom_point()


p <- ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State) +
    geom_point()
p
```

```{r}
p + 
  scale_colour_manual( values=c("blue","gray","red") ) + 
  labs(title = "Gene Expression Changes Upon Drug Treatment",
       x = " Control (no drug)",
       y= "Drug Treatment")
```

### 7. Going Further

```{r}
# install.packages("gapminder")
# install.packages("dplyr")
library(gapminder)
library(dplyr)

```

```{r}
gapminder_2007 <- gapminder |> 
  filter(year==2007)
```

```{r}
# Q. Complete the code below to produce a first basic scatter plot of this gapminder_2007 dataset:
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.5)

# color the points by the numeric variable population pop
ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, color = pop) +
  geom_point(alpha=0.8)

# adjust point size
ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, size = pop) +
  geom_point(alpha=0.5)
```

### 8. Bar Charts

```{r}
gapminder_top5 <- gapminder |>  
  filter(year==2007) |>  
  arrange(desc(pop)) |>  
  top_n(5, pop)

gapminder_top5
```

```{r}
# simple chart
ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = pop))
```

```{r}
# Create a bar chart showing the life expectancy of the five biggest countries by population in 2007.
ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = lifeExp))

```

```{r}
# Plot population size by country. Create a bar chart showing the population (in millions) of the five biggest countries by population in 2007.

ggplot(gapminder_top5) +
  aes(x=reorder(country, -pop), y=pop, fill=country) +
  geom_col()

```
