---
title: "Class17 HW: RNA-Seq Galaxy"
author: "Libby Gilmore (A69047570)"
format: pdf
toc: true
editor: visual
---

## Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(readr)
```


## Data Import

```{r}
galaxy_data <- read.table("expression_results.txt")
```


> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
galaxy_data |> 
  group_by(geno) |> 
  summarize(
    Sample_Size = n(),
    Median_Expression = median(exp)
  )
```


> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

The SNP seems to effect the expression of ORMDL3. Individuals with the A/A genotype show higher expression levels compared to those with the G/G genotype. This suggests that the presence of the A allele may be associated with increased expression of the ORMDL3 gene; or that the G allele could be linked to decreased expression. This could have additive effects if an individual is heterozygous (A/G), resulting in intermediate expression levels as seen below.
```{r}
ggplot(galaxy_data) +
  geom_boxplot(aes(x = geno, y = exp, fill = geno, notch=TRUE)) +
  labs(
    title = "Expression of ORMDL3 by Genotype",
    x = "Genotype",
    y = "Expression Level"
  ) +
  theme_minimal()
```

